---
title: "Problem set 2"
date: ""
format:
  html:
    toc: true
    number-sections: false
---
[Download source](https://raw.githubusercontent.com/ywanglab/STAT3000/refs/heads/main/psets/hw2_medium_hint.qmd)

For these exercises, do **not** load any packages other than **dslabs**.

Make sure to use **vectorization** whenever possible (avoid loops unless explicitly allowed).

```{r}
# Load only the allowed package
library(dslabs)
```

## 1

What is the sum of the first 150 positive integers? Use the functions `seq` and `sum` to compute the sum with R for any `n`.

```{R}
# Step 1: define n
n <- 150

# Step 2: create the sequence 1, 2, ..., n
x <- seq(1, n)

# Step 3: compute the sum of x
ans <- sum(x)

# Step 4: print ans
ans <- 11325
```


## 2

Load the `murders` dataset from **dslabs**. Use the function `str` to examine the structure of the `murders` object.

* What are the column names used by the data frame for these five variables: state name, abbreviation, region, population, total murders?
* Show the subset of `murders` showing states with **less than 1.2 per 100,000** deaths.
* Show **all** variables.

```{r}
# Examine structure
str(murders)

# Step: print the column names
 names(murders)
```
```{R}
# Step 1: compute murder rate per 100,000
rate <- murders$total / murders$population * 100000

# Step 2: create logical vector for rate < 1.2
idx <- rate < 1.2

# Step 3: subset murders
murders[idx, ]
```


## 3

Show the subset of `murders` showing states with **less than 1.2 per 100,000** deaths and in the **Northeast** of the US. Do **not** show the `region` variable.

```{r}
# Step 1: compute rate (or reuse from Q2)
# rate <- murders$total / murders$population * 100000

# Step 2: create logical vectors low and ne
# Step 2: create logical vectors
low <- rate < 1.2                   # low murder rate
ne <- murders$region == "Northeast" # Northeast region

# Step 3: combine conditions
keep <- low & ne

# Step 4: subset data and remove region column
out <- murders[keep, ]
out_no_region <- out[, names(out) != "region"]
out_no_region
```

## 4

Among states with a murder rate less than 1.2 per 100,000, show the **smallest population** state (show the state name, population, and rate).

```{r}
# Step 1: compute rate
rate <- murders$total / murders$population * 100000

# Step 2: restrict to low-rate states
low <- rate < 1.2

# Step 3: find the row index of the smallest population among those
i <- which(low)[which.min(murders$population[low])]

# Step 4: report as a small data frame
out_df <- data.frame(
  state = murders$state[i],
  population = murders$population[i],
  rate = rate[i]
)

# print the result
print(out_df)



```


## 5

Show the state with a population of **more than 8 million** with the **lowest** murder rate (show the state name, population, and rate).

```{r}
# Step 1: compute rate
rate <- murders$total / murders$population * 100000

# Step 2: restrict to states with population > 8 million
big <- murders$population > 8000000

# Step 3: find row index of smallest rate among those
i <- which(big)[which.min(rate[big])]

# Step 4: report as a small data frame
out_df <- data.frame(
  state = murders$state[i],
  population = murders$population[i],
  rate = rate[i]
)

# print the result
print(out_df)

```

## 6

Compute the murder rate for each **region** of the US (total murders divided by total population times 100,000). Return a data frame with one row per region and columns `region` and `rate`.

```{r}
# Step 1: total murders by region
murders_by_region <- tapply(murders$total, murders$region, sum)

# Step 2: total population by region
pop_by_region <- tapply(murders$population, murders$region, sum)

# Step 3: compute murder rates per 100,000
rate_by_region <- murders_by_region / pop_by_region * 100000

# Step 4: create data frame
region_rates <- data.frame(
  region = names(rate_by_region),
  rate = rate_by_region
)

# print the result
print(region_rates)

```

## 7

Create a vector of numbers that starts at 5, does not pass 60, and adds numbers in increments of 3/8.

How many numbers does the list have?

```{r}
# Step 1: define start, end, step
start <- 5
end <- 60
step <- 3/8

# Step 2: create sequence
v <- seq(from = start, to = end, by = step)

# Step 3: show first 6 values
head(v)

# Step 4: show length of the vector
length(v)

```

## 8

Make this data frame:

```{r}
temp_f <- c(72, 95, 41, 86, 78, 33)
city <- c("Chicago", "Lagos", "Oslo", "Rio de Janeiro", 
          "San Juan", "Toronto")
city_temps <- data.frame(name = city, temperature_f = temp_f)
city_temps
```

Add a new column called `temperature_c` containing the temperatures in Celsius. Keep all existing columns.

```{r}
# Step 1: create data frame
temp_f <- c(72, 95, 41, 86, 78, 33)
city <- c("Chicago", "Lagos", "Oslo", "Rio de Janeiro", "San Juan", "Toronto")

city_temps <- data.frame(
  name = city,
  temperature_f = temp_f
)

print(city_temps)

# Step 2: add Celsius column
city_temps$temperature_c <- (city_temps$temperature_f - 32) * 5/9

print(city_temps)

```

## 9

Write a function `euler2` that computes:

$$
S_n = 1 + \frac{1}{2^2} + \frac{1}{3^2} + \dots + \frac{1}{n^2}.
$$

Test your function at `n = 10` and `n = 100`.

```{r}
# Step 1: define the function
euler2 <- function(n) {
  k <- 1:n
  terms <- 1 / (k^2)
  sum(terms)
}

# Step 2: test the function
euler2(10)
euler2(100)

```

## 10

Plot $S_n$ versus $n$ for $n = 1,2,\dots,2000$ with a horizontal **dashed** line at $\pi^2/6$.

```{r}
# Step 1: create n-values
n_vals <- 1:2000

# Step 2: compute S_n for each n
S <- sapply(n_vals, euler2)

# Step 3: plot S_n vs n
plot(n_vals, S, type = "l", 
     xlab = "n", ylab = expression(S[n]))

# Step 4: add dashed line at pi^2/6
abline(h = pi^2 / 6, lty = 2)

```

## 11

Use `%in%` and `state.abb` to create a logical vector for: AL, AK, AZ, AR, AA.

```{r}
test <- c("AL", "AK", "AZ", "AR", "AA")

# Step 1: check which are real state abbreviations
is_real <- test %in% state.abb

# Step 2: print
is_real

```

## 12

Report the one entry that is **not** an actual abbreviation.

```{r}
test <- c("AL", "AK", "AZ", "AR", "AA")

# Step 1: logical vector for not real
not_real <- !(test %in% state.abb)

# Step 2: find index
idx <- which(not_real)

# Step 3: print the entry
test[idx]

```

## 13

Using `%in%`, show all variables for Florida, California, and New York, **in that order**.

```{r}
targets <- c("Florida", "California", "New York")

# Step 1: subset murders to those states
sub <- murders[murders$state %in% targets, ]

# Step 2: reorder rows using match()
sub_ordered <- sub[match(targets, sub$state), ]

# Step 3: print
sub_ordered

```

## 14

Write a function `vander_helper(x, n)` that returns $(1, x, x^2, \dots, x^n)$. Show results for `x=2`, `n=6`.

Restrictions: no loop.

```{r}
vander_helper <- function(x, n) {
  x^(0:n)
}

# Test
vander_helper(2, 6)

```

## 15

Create a vector using:

```{r}
n <- 20000
p <- 0.35
set.seed(2025-9-18)
x <- sample(c(0,1), n, prob = c(1 - p, p), replace = TRUE)
```

Compute the length of each **stretch of consecutive 1s** (run lengths of 1s) and plot the distribution.

* Do **not** use a loop.
* Hint: use `rle(x)`.

Then compare empirical proportions to the geometric prediction for run lengths 1 through 8.

```{r}
# Step 1: create vector
n <- 20000
p <- 0.35
set.seed(2025-9-18)
x <- sample(c(0,1), n, prob = c(1 - p, p), replace = TRUE)

# Step 2: run lengths
r <- rle(x)
ones_lengths <- r$lengths[r$values == 1]

# Step 3: plot distribution
hist(ones_lengths, breaks = 30, main = "Distribution of Run Lengths of 1s",
     xlab = "Run Length", ylab = "Frequency")

# Step 4: empirical proportions for k=1:8
tab <- table(ones_lengths)
emp_counts <- as.numeric(tab[as.character(1:8)])
emp_counts[is.na(emp_counts)] <- 0
emp_probs <- emp_counts / length(ones_lengths)

# Step 5: theoretical probabilities
k <- 1:8
theory_probs <- (1 - p) * p^(k - 1)

# Step 6: comparison
comparison <- data.frame(
  run_length = k,
  empirical_prob = emp_probs,
  theory_prob = theory_probs
)

comparison

```

## 16

In the `murders` dataset:

1. Compute the national average murder rate.
2. Create labels using `ifelse`:

   * `"High Crime, High Pop"` if rate > national average and pop > 6 million
   * `"High Crime, Low Pop"` if rate > national average and pop ≤ 6 million
   * `"Lower Crime"` otherwise

Then show a `table()` of the labels.

```{r}
# Step 1: state-level rates
rate <- murders$total / murders$population * 100000

# Step 2: national average rate
national_rate <- sum(murders$total) / sum(murders$population) * 100000

# Step 3: logical vectors
high_crime <- rate > national_rate
high_pop <- murders$population > 6000000

# Step 4: create labels
labels <- ifelse(high_crime & high_pop, "High Crime, High Pop",
                 ifelse(high_crime & !high_pop, "High Crime, Low Pop",
                        "Lower Crime"))

# Step 5: show table
table(labels)

```

## 17

What is the murder rate of the state that ranks **12th** in terms of murder rate (from highest to lowest)?

Show your work using `order` (and optionally check with `sort` or `rank`).

```{r}
# Step 1: rate vector
rate <- murders$total / murders$population * 100000

# Step 2: ord <- order(rate, decreasing = TRUE)
ord <- order(rate, decreasing = TRUE)

# Step 3: i <- ord[12]
i <- ord[12]

# Step 4: report state and rate
data.frame(state = murders$state[i], rate = rate[i])

```

## 18

Write a function `compute_harmonic_mean` that returns the harmonic mean of a numeric vector, but returns `NA` if any values are zero or negative. Test on `c(1,2,4,8)` and show it is about `2.133333`.

```{r}
compute_harmonic_mean <- function(x) {
  if(any(x <= 0)) return(NA)
  n <- length(x)
  n / sum(1 / x)
}

# Test
compute_harmonic_mean(c(1, 2, 4, 8))

```

## 19

Create a function `safe_divide(x, y)` that returns `x/y` but returns `"Cannot divide by zero"` when `y` is zero. Make it work element-wise on vectors (vectorized). Test it on:

```r
x <- c(10, 20, 30)
y <- c(2, 0, 5)
```

```{r}
safe_divide <- function(x, y) {
  out <- x / y
  out <- as.character(out)
  out[y == 0] <- "Cannot divide by zero"
  out
}

# Test
x <- c(10, 20, 30)
y <- c(2, 0, 5)
safe_divide(x, y)

```

## 20

Write a function `classify_state_safety(state_name)` that returns:

* `"Very Safe"` if rate < 1
* `"Safe"` if 1 ≤ rate < 3
* `"Moderate"` if 3 ≤ rate < 5
* `"High Risk"` if rate ≥ 5
* `"State not found"` if the state is not in the dataset

Test on `"Vermont"`, `"Texas"`, `"California"`, `"NotAState"`.

Then use `sapply` to classify all states and use `table()` to count how many fall into each category.

```{r}
# Step 1: compute a named vector of rates
rate <- murders$total / murders$population * 100000
rate_named <- setNames(rate, murders$state)

classify_state_safety <- function(state_name) {
  # Step 2: check if state_name is in names(rate_named)
  if(!(state_name %in% names(rate_named))) {
    return("State not found")
  }
  
  # Step 3: pull out r <- rate_named[state_name]
  r <- rate_named[state_name]
  
  # Step 4: return the correct label using if/else
  if(r < 1) {
    return("Very Safe")
  } else if(r < 3) {
    return("Safe")
  } else if(r < 5) {
    return("Moderate")
  } else {
    return("High Risk")
  }
}

# Tests:
# classify_state_safety("Vermont")
# classify_state_safety("Texas")
# classify_state_safety("California")
# classify_state_safety("NotAState")

# Step 5: classify all states with sapply
cats <- sapply(murders$state, classify_state_safety)

# Step 6: count categories
table(cats)

```


## Convert to  a PDF file
In a (Linux terminal), run the following command (install any missing packages on the fly)

```
# change path_to_hw2.qmd to something like hw/hw2.qmd if the current directory is a parent diretory of hw/
quarto render hw/HW2.qmd --to pdf --output-dir docs
```





